Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-20 talk/iwtc11/paper.tex
explain that allocation removal leads to type specialization
Merge the CSE and heap optimization sections to save space, since they say mostly the same thing. From the trace, only remove one of the gets. This makes it easier to explain, and the other one is removed by allocation removal anyway

Several benchmarks, with few guard failures, executed on the
PyPy python JIT show over 2 
times increase in speed when loop peeling was introduced. This makes
some of them almost match optimized C performance and become over XXX
times faster than cpython.
\end{abstract}

@@

Note that the guard on $p_5$ is removed even though $p_5$ is not loop
invariant, which shows that loop invariant code motion is not the only
effect of loop peeling.

\subsection{Heap Caching}

XXX gcc calls this store-sinking and I'm sure there are some
references in the literature (none at hand though). This is a ``typical''
compiler optimization.

The objective of heap caching is to remove \lstinline{get} and
\lstinline{set} operations whose results can be deduced from previous
\lstinline{get} and \lstinline{set} operations. Exact details of the
process are outside the scope of this paper. We only consider the interaction
with loop peeling.

The issue at hand is to keep the peeled loop a proper
trace. Consider the \lstinline{get} operation on line 19 of
Figure~\ref{fig:peeled-trace}. The result of this operation can be
deduced to be $i_4$ from the \lstinline{set} operation on line
12. Also, the result of the \lstinline{get} operation on line 22 can
be deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 19 and 22 from the trace and
replace $i_6$ with $i_4$ and $i_7$ with $i_3$.

After that, the peeled loop
will no longer be in SSA form as it operates on $i_3$ and $i_4$
which are not part of it. The solution is to extend the input
arguments, $J$, with those two variables. This will also extend the
jump arguments of the preamble, which is also $J$. 
Implicitly that also extends the jump arguments of the peeled loop, $K$,
since they are the image of $J$ under $m$. For the example $I$ has to
be replaced by $\hat I$ which is formed as a concatenation of $I$ and
$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
$\hat K$ which is formed as a concatenation of $K$ and 
$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$. XXX: Maybe we should
replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?

In general what is needed is for the heap optimizer is to keep track of
which variables from the preamble it reuses in the peeled loop.
It has to construct a vector, $H$,  of such variables which
can be used to update the input and jump arguments using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:peeled-trace} will be optimized to:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($L_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)

$L_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
# inside f: y = y.add(step)
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{8}$ = $i_{4}+i_{3}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($L_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
\end{lstlisting}

Note how the loop invaraint \lstinline{get} on $p_0$ was moved out of
the loop, and how the non loop invariant \lstinline{get} on $p_5$ was
removed entierly.

\subsection{Common Subexpression Elimination}
If a pure operation appears more than once in the trace with same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. When that is combined with loop
peeling, the single execution of the operation is placed in the
preamble. That is, loop invariant pure operations are moved out of the
loop. The interactions here are the same as in the previous
section. That is, a vector, $H$, of variables produced in the preamble
and used in the peeled loop needs to be constructed. Then the jump and
input arguments are updated according to
Equation~\ref{eq:heap-inputargs} and Equation~\ref{eq:heap-jumpargs}.

@@

@@

\subsection{Python}
The python interpreter of the PyPy framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. In Python the algorithm is implemented
  on top of a custom image class that is specially designed for the
  problem. It ensures that there will be no failing guards, and makes
  a lot of the two dimension index calculations loop invariant. The
  intention here is twofold. It shows that the performance-impact of
  having wrapper classes giving objects some application-specific
  properties can be negligible. This is due to the inlining performed
  during the tracing and the allocation removal of the index objects
  introduced. It also shows that it is possible to do some low-level
  hand optimizations of the Python code and hide those optimization
  under a nice interface without loosing performance.
\end{itemize}

>>>>

Several benchmarks, with few guard failures, executed on the
PyPy Python JIT show over 2
times increase in speed when loop peeling was introduced. This makes
some of them almost match optimized C performance and become over XXX
times faster than CPython.
\end{abstract}

@@

Note that the guard on $p_5$ is removed even though $p_5$ is not loop
invariant, which shows that loop invariant code motion is not the only
effect of loop peeling. Loop peeling can also remove guards that are implied by
the guards of the previous iteration.

\subsection{Common Subexpression Elimination and Heap Optimizations}

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. PyPy's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value\footnote{We perform a simple type-based alias analysis to know which
writes can affect which reads. In addition writes on newly allocated objects
can never change the value of old existing ones.}.

When that is combined with loop peeling, the single execution of the operation
is placed in the preamble. That is, loop invariant pure operations and heap
reads are moved out of the loop.

Consider the \lstinline{get} operation on line 22 of
Figure~\ref{fig:peeled-trace}. The result of this operation can be
deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 22 from the trace and
replace $i_7$ with $i_3$. Afterwards the trace is no longer in the correct
form, because the argument $i_3$ is not passed along the loop arguments. It
thus needs to be added there.

The trace from Figure~\ref{fig:peeled-trace} will therefore be optimized to:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($L_1$, $p_{0}$, $p_{5}$, $i_3$)

$L_1$($p_{0}$, $p_{5}$, $i_3$):
# inside f: y = y.add(step)
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{8}$ = $i_{4}+i_{3}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($L_1$, $p_{0}$, $p_{9}$, $i_3$)
\end{lstlisting}

In general, after loop peeling and redundant operation removal the peeled loop
will no longer be in SSA form as it operates on variables that are the result
of pure operations in the preamble. The solution is to extend the input
arguments, $J$, with those variables. This will also extend the
jump arguments of the preamble, which is also $J$. 
Implicitly that also extends the jump arguments of the peeled loop, $K$,
since they are the image of $J$ under $m$. For the example $I$ has to
be replaced by $\hat I$ which is formed by appending $i_3$ to $I$.
At the same time $K$ has to be replaced by
$\hat K$ which is formed by appending $m\left(i_3\right)=i_7$ to $K$.
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$.

In general what is needed is to keep track of
which variables from the preamble it reuses in the peeled loop.
It has to construct a vector, $H$,  of such variables which
can be used to update the input and jump arguments using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$.

@@

XXX explain that this is effectively type-specializing a loop

@@

\subsection{Python}
The Python interpreter of the PyPy framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. In Python the algorithm is implemented
  on top of a custom image class that is specially designed for the
  problem. It ensures that there will be no failing guards, and makes
  a lot of the two dimension index calculations loop invariant. The
  intention here is twofold. It shows that the performance-impact of
  having wrapper classes giving objects some application-specific
  properties can be negligible. This is due to the inlining performed
  during the tracing and the allocation removal of the index objects
  introduced. It also shows that it is possible to do some low-level
  hand optimizations of the Python code and hide those optimization
  under a nice interface without loosing performance.
\end{itemize}
